---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{R, echo = FALSE, results = "hide"}
#The following libraries will be requried for this data analysis
library(lubridate)
library(ggplot2)
library(dplyr)
```

```{R}
unzip("activity.zip", overwrite = TRUE) #unzip the dataset
repdata <- read.csv("activity.csv") #read the dataset
repdata <- group_by(repdata, date) #Group the dataset by date
```

## What is mean total number of steps taken per day?
```{r}
#Calculate the total number of steps taken per day
perday <- summarize(repdata, steps = sum(steps))

#Histogram of total number of steps taken per day
hist(perday$steps, xlab = "Number of steps per day", main = "Histogram of Steps taken per day")

#Mean and median of total number of steps taken per day
day_sum <- summarize(repdata, mean_steps = mean(steps, na.rm = T), median_steps = median(steps, na.rm = T))
print.data.frame(day_sum)
```

## What is the average daily activity pattern?
```{r}
#Calculate and insert the interval number as a column
repdata$int_num <- 0 #create a new column

#Put incremental numbers in int_num column and reset every time the date changes
for(i in 2:17568){
  if(repdata[i,2] == repdata[i-1,2]){
    repdata[i,4] = repdata[i-1,4]+1
  } else{
    repdata[i,4] = 1
  }
}

#Drawing the time series plot
repdata <- group_by(repdata, int_num)
int_sum <- summarize(repdata, steps = mean(steps, na.rm = T))
with(int_sum, plot(int_num, steps, type = "l", xlab = "Interval number", ylab = "Number of steps", main = "Time series plot of average daily activity pattern"))

#Interval where number of steps is maximum
max_steps <- int_sum[int_sum$steps == max(int_sum$steps, na.rm = TRUE),]
max_steps <- max_steps[2,]
```

The interval with maximum steps is `r max_steps`.

## Inputing missing values
```{r, results = "hide"}
#Total number of missing values in the dataset
missing <- sum(is.na(repdata$steps))
```

The number of rows with missing data is `r missing`.

```{r}
#Create a new dataset which is a copy of the original dataset
repdata2 <- repdata

#Fill in the missing values using the mean of the particular interval
repdata2[is.na(repdata2$steps),]$steps <- 
  round(int_sum$steps[match(repdata2[is.na(repdata2$steps),]$int_num, int_sum$int_num)])

repdata2 <- group_by(repdata2, date)
perday2 <- summarize(repdata2, steps = sum(steps))

#Histogram and mean and median of per day steps
hist(perday2$steps, xlab = "Number of steps per day", main = "Histogram of Steps taken per day")
day_sum2 <- summarize(repdata2, mean_steps = mean(steps, na.rm = T), median_steps = median(steps, na.rm = T))
print.data.frame(day_sum2)
```

###Comparing the histograms of data with and without missing values

```{r, fig.height = 8}
#Comparison of plots with a without missing values
par(mfcol = c(2,1))
hist(perday$steps, xlab = "Number of steps per day", main = "Histogram of Steps taken per day (with missing values)")
hist(perday2$steps, xlab = "Number of steps per day", main = "Histogram of Steps taken per day (Without missing values)")
```

*Second and third columns are slightly higher*

## Are there differences in activity patterns between weekdays and weekends?
```{r}
#Define a new data frame dividing days into weekday and weekend
weekend <- data.frame(day = c("Sunday","Monday","Tuesday", "Wednesday","Thursday","Friday","Saturday"), type = c("Weekend","Weekday","Weekday","Weekday","Weekday","Weekday","Weekend"))

#Make sure that date column is of type POSIXct
repdata2$date <- ymd(repdata2$date)

#Add a column extracting the day from the date
repdata2 <- mutate(repdata2, day = weekdays(date))

#match with weekend data frame to label weekdays and weekends
repdata2$day_type <- weekend$type[match(repdata2$day, weekend$day)]
rm(weekend) #remove the weekend dataframe as it is no longer useful

#Summarize the data set based on weekdays and weekends
repdata2<- group_by(repdata2, int_num, day_type)
step_summ <- summarize(repdata2, mean_steps = mean(steps))

#Make panel plot of weekday Vs. Weekend
qplot(int_num, mean_steps, data = step_summ, geom = "line", facets = day_type~., xlab = "Interval Numbers", ylab = "Average number of steps")
```

*During weekends, the number of steps taken is more evenly distributed throughout the day*

